# R Interface Examples

The R package `genieclust` (**TODO**) *will* be available on CRAN.
To install it, call:

```{r install, eval=FALSE}
# TODO: install.packages("genieclust")
devtools::install_github("gagolews/genieclust")
```


Below are a few basic examples on how to interact with the package
(partially based on Marek's [forthcoming book](https://lmlcr.gagolewski.com)).


```{r load}
library("genieclust")
```

```{r hello,echo=FALSE}
# library("reticulate")
# use_python("/usr/bin/python3")
library("knitr")
options(width=1000)

opts_chunk$set(
    fig.path='figures_r/',
    cache.path='cache_r/',
    autodep=FALSE,
    cache=TRUE,
    fig.width=10,
    fig.height=4,
    dpi=300,
    dev="CairoPNG",
    dev.args=list(bg="transparent",pointsize=12)
)

source("graphics_hacks.R")
set.seed(123)
```


Let's consider the Sustainable Society Indices Dataset
(see http://www.ssfindex.com/)
that measures the Human, Environmental,  and Economic Wellbeing
in each country based on 7 categories on the scale $[0, 10]$.

```{r ssi-load}
# see https://github.com/gagolews/genieclust/tree/master/devel/sphinx/rmd
ssi <- read.csv("ssi_2016_categories.csv", comment.char="#")
X <- as.matrix(ssi[,-1])    # everything except the Country (1st) column
dimnames(X)[[1]] <- ssi[,1] # set row names
head(X)  # preview
```


`genieclust` has an interface compatible with the R's workhorse
for hierarchical clustering, `stats::hclust()`.
Yet, for efficiency reasons, it's better to feed `genieclust::glust()`
with the input matrix directly:


```{r dist-gclust}
# faster than gclust(dist(X)):
h <- gclust(X)  # default: gini_threshold=0.3, distance="euclidean"
print(h)
```

In order to extract a desired *k*-partition, we can call `stats::cutree()`:

```{r cutree}
y_pred <- cutree(h, k=3)
sample(y_pred, 25)  # preview
```

This gives the cluster ID allocated to each country.
Let's depict the obtained partition with the `rworldmap` package:

```{r ssi-map,echo=-c(1,2,7),results="hide",message=FALSE,fig.cap="Countries grouped w.r.t. the SSI categories."}
par(ann=FALSE)  # no axes
par(mar=c(0, 0, 0, 0))  # no figure margins
library("rworldmap")  # see the package's manual for details
mapdata <- data.frame(Country=dimnames(X)[[1]], Cluster=y_pred)
mapdata <- joinCountryData2Map(mapdata, joinCode="NAME", nameJoinColumn="Country")
mapCountryData(mapdata, nameColumnToPlot="Cluster", catMethod="categorical",
    missingCountryCol="white", colourPalette=palette.colors(3, "Okabe-Ito"),
    mapTitle="")
```


We can compute, e.g., the average indicators in each identified group:

```{r ssi-colmeans}
t(aggregate(as.data.frame(X), list(Cluster=y_pred), mean))[-1, ]
```


Plotting of dendrograms is also possible.
For greater readability, we'll restrict ourselves to a smaller sample, say,
to the 37 members of the [OECD](https://en.wikipedia.org/wiki/OECD):

```{r ssi-oecd}
oecd <- c("Australia", "Austria", "Belgium", "Canada", "Chile", "Colombia",
"Czech Republic", "Denmark", "Estonia", "Finland", "France", "Germany",
"Greece", "Hungary", "Iceland", "Ireland", "Israel", "Italy", "Japan",
"Korea, South", "Latvia", "Lithuania", "Luxembourg", "Mexico", "Netherlands",
"New Zealand", "Norway", "Poland", "Portugal", "Slovak Republic", "Slovenia",
"Spain", "Sweden", "Switzerland", "Turkey", "United Kingdom", "United States")
X_oecd <- X[dimnames(X)[[1]] %in% oecd, ]
```


```{r ssi-oecd-dendrogram, echo=-1, fig.cap="Cluster dendrogram for the OECD countries."}
par(mar=c(0.5, 2, 0.5, 0.5))
h_oecd <- gclust(X_oecd)
plot(h_oecd)
```




Final notes:

* If we are only interested in a specific partition,
calling `genie()` directly will be a little faster than referring to
`cutree(gclust(...))`.

* `genieclust` also features partition similarity scores
(such as the Adjusted Rand or the Pair Sets Index) that can be used as
external cluster validity measures.

For more details, refer to the package's documentation.
Don't forget to check out the Python examples regarding noise point detection,
benchmarking, timing, etc.






