# Basic Usage (R)

The latest stable release of the R package **genieclust** is available from the
[CRAN](https://CRAN.R-project.org/package=genieclust) repository.
Install it by calling:

```{r install, eval=FALSE}
install.packages("genieclust")
```


Below are a few basic examples of interacting with the package.

```{r load}
library("genieclust")
```


Let's take the [Sustainable Society Indices](http://www.ssfindex.com/)
dataset that measures the Human, Environmental, and Economic Wellbeing
in each country. There are seven categories on the scale $[0, 10]$.

```{r ssi-load}
# see https://github.com/gagolews/genieclust/tree/master/devel/sphinx/weave
ssi <- read.csv("ssi_2016_categories.csv", comment.char="#")
X <- as.matrix(ssi[,-1])    # everything except the Country (1st) column
dimnames(X)[[1]] <- ssi[,1] # set row names
head(X)  # preview
```


The API of **genieclust** is compatible with R's workhorse
for hierarchical clustering, `stats::hclust()`, which accepts
a complete pairwise distance matrix.  Yet, for better performance,
it is better to feed `genieclust::gclust()` with the input data matrix
directly:


```{r dist-gclust}
# faster than gclust(dist(X)):
h <- gclust(X)  # default: gini_threshold=0.3, distance="euclidean"
print(h)
```

In order to extract a desired *k*-partition, we can call `stats::cutree()`:

```{r cutree}
y_pred <- cutree(h, k=3)  # also: genie(X, 3)
sample(y_pred, 25)  # preview
```

This gives the cluster IDs allocated to each country.


::::{note}
If we are only interested in the partition of a specific cardinality,
calling `genie()` directly will be slightly faster than referring to
`cutree(gclust(...))`.
::::

Let's depict the obtained partition using the **rworldmap** package:

```{r ssi-map,echo=-c(1,2,7),results="hide",message=FALSE,fig.cap="Countries grouped with respect to the SSI categories"}
par(ann=FALSE)  # no axes
par(mar=c(0, 0, 0, 0))  # no figure margins
library("rworldmap")  # see the package's manual for details
mapdata <- data.frame(Country=dimnames(X)[[1]], Cluster=y_pred)
mapdata <- joinCountryData2Map(mapdata, joinCode="NAME", nameJoinColumn="Country")
mapCountryData(mapdata, nameColumnToPlot="Cluster", catMethod="categorical",
    missingCountryCol="white", colourPalette=palette.colors(3, "R4"),
    mapTitle="")
```


We can compute, for instance, the average indicators in each identified group:

```{r ssi-colmeans}
t(aggregate(as.data.frame(X), list(Cluster=y_pred), mean))[-1, ]
```


## Dendrograms

Dendrogram plotting is also possible.
For better readability, we will restrict ourselves to a smaller sample;
namely, to the 37 members of the [OECD](https://en.wikipedia.org/wiki/OECD):

```{r ssi-oecd}
oecd <- c("Australia", "Austria", "Belgium", "Canada", "Chile", "Colombia",
"Czech Republic", "Denmark", "Estonia", "Finland", "France", "Germany",
"Greece", "Hungary", "Iceland", "Ireland", "Israel", "Italy", "Japan",
"Korea, South", "Latvia", "Lithuania", "Luxembourg", "Mexico", "Netherlands",
"New Zealand", "Norway", "Poland", "Portugal", "Slovak Republic", "Slovenia",
"Spain", "Sweden", "Switzerland", "Turkey", "United Kingdom", "United States")
X_oecd <- X[dimnames(X)[[1]] %in% oecd, ]
```


```{r ssi-oecd-dendrogram, echo=-1, fig.height=5.9375, fig.cap="Cluster dendrogram for the OECD countries"}
par(mar=c(2, 2, 1, 6))
h_oecd <- gclust(X_oecd)
plot(as.dendrogram(h_oecd), horiz=TRUE)
```


## Outlier Detection with Deadwood

The Deadwood outlier detection algorithm can be run on the clustered dataset
to identify anomalous points in each cluster.
See the [**deadwood**](https://deadwood.gagolewski.com/) package
tutorials for more details.


## Remarks

**genieclust** also features partition similarity scores
(such as the adjusted Rand index) and internal cluster validity measures
(e.g., the Caliński-Harabasz index).

For more details, refer to the package's {any}`documentation <../rapi>`.

To learn more about R, check out Marek's open-access textbook
[Deep R Programming](https://deepr.gagolewski.com/) {cite}`deepr`.
